How to Avoid OOTA Without Really Trying

### 2025-03-05 paper/ootavacuous.tex
ootavacuous: Clarify departure from prior OOTA work
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Some workers recommend avoiding OOTA by forcing prior relaxed
loads to be ordered before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
but this can require executing real
instructions~\cite[Section 7.1]{Maranget2012TutorialARMPower},
consuming real time and real electrical power to solve a strictly
theoretical problem.
% This might have been acceptable in the 1960s of some of the authors'
% youths, but it is now the year 2024.
Furthermore, as discussed in
Section~\ref{sec:Global Optimization Can Destroy Dependencies},
cross-thread optimizations can defeat this ordering constraint.

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been accepted by C++ implementers.
As of 2024, Mark Batty is working on modular relaxed dependencies,
which may have the added ability of checking compiler
optimizations~\cite{JayRichards2024SymbolicMrder}.

@@

All this work focused on either identifying OOTA or helping C++
implementations to avoid it.
None applied real-world hardware ordering constraints to the problem
of avoiding OOTA cycles.
This paper's contribution is to show how these constraints prevent
real-world C++ implementations from exhibiting OOTA behavior.

>>>>

Some researchers recommend avoiding OOTA by forcing prior relaxed
loads to be ordered before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352,PiezhaoOu2018OOTApapl},
but this can require executing real
instructions~\cite[Section 7.1]{Maranget2012TutorialARMPower},
consuming real time and real electrical power to solve a strictly
theoretical problem.
% This might have been acceptable in the 1960s of some of the authors'
% youths, but it is now the year 2024.
Furthermore, as discussed in
Section~\ref{sec:Global Optimization Can Destroy Dependencies},
cross-thread optimizations can defeat this ordering constraint.

Other researchers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been accepted by C++ implementers.
As of 2024, Mark Batty is working on modular relaxed dependencies,
which may have the added ability of checking compiler
optimizations~\cite{JayRichards2024SymbolicMrder}.

@@

All this work focused on either identifying OOTA or helping C++
implementations to avoid it.
None applied real-world hardware ordering constraints (based on
long-standing laws of physics) to the problem of avoiding OOTA cycles.

Augmenting C++ abstractions with these constraints would of course
limit applicability of this paper's results to C++ programs running
on real-world computer systems.
However, a quarter century of OOTA research based on pure C++ abstractions
has resulted in incremental progress at best, and real-world computer
systems comprise an exceedingly important part of the overall OOTA
problem.
In the wise words of Peter J.~Denning, ``Maybe we need to occasionally
descend from the high clouds of our abstractions to the concrete earthy
concerns of everyday life.''~\cite{PeterJDenning2025Abstraction}.

Therefore, this paper's contribution is to show how these constraints
prevent real-world C++ implementations from exhibiting OOTA behavior,
and not with minimal cost~\cite{PiezhaoOu2018OOTApapl}, but rather with
no cost at all.
